Simulating High-Dimensional Multivariate Data

### 2020-11-12 010-background.Rmd
3rd draft. Finalizing pdf output and small issues

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $\rho_s$ for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

>>>>

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

### 2020-11-12 020-simulation-algorithm.Rmd
3rd draft. Finalizing pdf output and small issues

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
In general, there is no closed form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
For low-dimensional problems, methods and tools exist to match Pearson correlations precisely (see @Chen2001; @Xia17; @MB13), including the publicly available `nortaRA` R package.

@@

Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution (@Song00) in many practical applications.
The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evalutations]({#simulations}).

@@

The pre-processing (Step1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).

>>>>

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
In general, there is no closed form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
For low-dimensional problems, methods and tools exist to match Pearson correlations precisely, see @Chen2001; @Xia17; @MB13 and the publicly available `nortaRA` R package.

@@

Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evalutations]({#simulations}).

@@

The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).

### 2020-11-12 050-examples.Rmd
3rd draft. Finalizing pdf output and small issues

```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
## knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
# reticulate::use_condaenv("bigsimr")
library(bigsimr)
library(tidyverse)
cores <- as.integer( parallel::detectCores() - 1 )
set.seed(10142020)
```

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataFig).
Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
exampleGenes <- readRDS(file = './data/exampleGenes.rds')
GGally::ggpairs(
  data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
  upper = list(
    continuous = GGally::wrap('cor', method = "spearman")
  ), 
) + theme_bw()
```

@@

```{r ch050-figBRCA, echo=F, eval=T, cache=T, fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
## estimate rho from sims and plot
nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
## triplot
par(mfrow=c(1,3))
plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
     xlim = c(-1, 1), ylim = c(-1, 1),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
     xlab = "Target Correlation", ylab = "Estimated Correlation",
     main = "Spearman Cor. Matching")
abline(a = 0, b = 1, col="red", lty="dashed")
## estimate prob,size from sims and plot
nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
plot(sizes, nbinom_size_prob_hat[1,],
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Size", ylab = "Estimated Size",
     main = "Marginal NB Size Matching")
abline(a=0, b=1, col="red", lty="dashed")
plot(log(probs), log(nbinom_size_prob_hat[2,]),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
     main = "Marginal NB Prob Matching")
abline(a=0, b=1, col="red", lty="dashed")
par(mfrow=c(1,1))
```

@@

where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$).

>>>>

```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
## knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
# reticulate::use_condaenv("bigsimr")
library(bigsimr)
library(tidyverse)
cores <- as.integer( parallel::detectCores() - 1 )
set.seed(10142020)
```

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.', dpi=36}
exampleGenes <- readRDS(file = './data/exampleGenes.rds')
GGally::ggpairs(
  data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
  upper = list(
    continuous = GGally::wrap('cor', method = "spearman")
  ), 
) + theme_bw()
```

@@

```{r ch050-figBRCA, echo=F, eval=T, cache=T,fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).", dpi=36}
## estimate rho from sims and plot
nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
## triplot
par(mfrow=c(1,3))
plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
     xlim = c(-1, 1), ylim = c(-1, 1),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
     xlab = "Target Correlation", ylab = "Estimated Correlation",
     main = "Spearman Cor. Matching")
abline(a = 0, b = 1, col="red", lty="dashed")
## estimate prob,size from sims and plot
nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
plot(sizes, nbinom_size_prob_hat[1,],
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Size", ylab = "Estimated Size",
     main = "Marginal NB Size Matching")
abline(a=0, b=1, col="red", lty="dashed")
plot(log(probs), log(nbinom_size_prob_hat[2,]),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
     main = "Marginal NB Prob Matching")
abline(a=0, b=1, col="red", lty="dashed")
par(mfrow=c(1,1))
```

@@

where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $0.1708$.

### 2020-11-12 index.Rmd
3rd draft. Finalizing pdf output and small issues

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author: "Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler"
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Anna K. Panorska
  - Juli Petereit
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---
